---
name: prologue
---



---

## Statistics Inform Policy

[Policy:]{.hi} In 2017, the University of Oregon started requiring first-year students to live on campus.

. . .

[Rationale:]{.hi} First-year students who live on campus fare better:

- _[80 percent]{.hii} more likely_ to graduate in four years.
- Second-year retention rate _[5 percentage points higher]{.hii}_.
- GPAs _[0.13 points higher]{.hii}_, on average. 

. . .

_Do these comparisons suggest the policy improves student outcomes?_

. . .

_Do they describe the effect of living on campus?_

. . .

_Do they describe something else?_

---

## Other Things Equal

The UO's interpretation of those comparisons warrants skepticism.

- The decision to live on campus is probably related to family wealth and interest in school.
- Family wealth and interest in school are related to academic achievement.

. . .

[Why might I be worried?]{.hi} 

: The difference in outcomes between those on and off campus is not an [_all else equal_]{.note} comparison.

[Upshot:]{.hi} Can't attribute the difference in outcomes solely to living on campus.


---

Living on campus ([D]{.hi}) increases student welfare ([Y]{.hi})

```{r}
#| label: dag-ex-setup
#| echo: FALSE
#| include: FALSE



# The full DAG
dag_full = dagify(
  Y ~ D,
  Y ~ W,
  D ~ W,
  coords = tibble(
    name = c("Y", "D", "W"),
    x = c(1, 3, 2),
    y = c(2, 2, 1)
  )
)
# Convert to data.table
dag_dt = dag_full %>% fortify() %>% setDT()
# Add indicators for paths
dag_dt[, `:=`(
  path1 = (name == "D" & to == "Y") | (name == "Y"),
  path2 = (name == "D" & to == "W") | (name == "W" & to == "Y") | (name == "Y")
)]
# Shorten segments
mult = 0.15
dag_dt[, `:=`(
  xa = x + (xend-x) * (mult),
  ya = y + (yend-y) * (mult),
  xb = x + (xend-x) * (1-mult),
  yb = y + (yend-y) * (1-mult)
)]
```

```{r}
#| label: dag-ex-ovb-00
#| echo: FALSE
#| fig.height: 3
#| fig.width: 6

# Filter data to include only D and Y nodes and the edge between them
dag_dt_filtered <- dag_dt[((name == "D" & to == "Y") | (name %in% c("D", "Y"))),]

# Plot the filtered DAG
ggplot(
  data = dag_dt_filtered,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = hi,
  shape = 21,
  stroke = 0.6
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = hi,
  size = 1.2,
  lineend = "round"
) +
geom_text(
  data = dag_dt_filtered[, .(name, x, y, xend = x, yend = y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = hi,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[, min(x)] * 0.95, dag_dt[, max(x)] * 1.05),
  ylim = c(dag_dt[, min(y)] * 0.8, dag_dt[, max(y)] * 1.1)
)

```

<br>
<br>

---

Living on campus ([D]{.hi}) increases student welfare ([Y]{.hi})


```{r}
#| label: dag-ex-ovb-01
#| echo: FALSE
#| fig.height: 3
#| fig.width: 6

# Plot the full DAG
ggplot(
  data = dag_dt,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = hi,
  shape = 21,
  stroke = 0.6
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = hi,
  size = 1.2,
  lineend = "round"
) +
geom_text(
  data = dag_dt[, .(name, x, y, xend = x, yend = y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = hi,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[, min(x)] * 0.95, dag_dt[, max(x)] * 1.05),
  ylim = c(dag_dt[, min(y)] * 0.8, dag_dt[, max(y)] * 1.1)
)
```

But parental income ([W]{.hi}) impacts both student welfare ([Y]{.hi}) and living on campus ([D]{.hi}). [Failing to account for parental income will bias comparisons]{.fragment}

---

## All else equal

> _Ceteris paribus_, _all else held constant_, _etc_.

<br>

. . .

#### [A high bar]{.hi}

When [all else equal]{.note}, statistical comparisons detect causal relationships.

. . .

(Micro)economics has developed a comparative advantage in understanding where [_all else equal_]{.note} comparisons can and cannot be made.

- Anyone can retort "_correlation doesn't necessarily imply causation_."
- Understanding _why_ is difficult, but useful for learning from data.
