---
name: fundamental-problem
---

---

## Causal identification

[Goal:]{.note} 

: Identify the effect of a treatment, $D_i$, on individual $i$'s outcome, $Y_{D,i}$.

. . .

<br>

Ideally, we could calculate the [treatment effect]{.hi} *for each individual* $i$ as

$$Y_{1,i} - Y_{0,i}$$ 

- $Y_{1,i}$: the outcome for person $i$ when she receives the treatment.
- $Y_{0,i}$: the outcome for person $i$ when she doesn't receive the treatment.

. . .

Referred to as [potential outcomes]{.note}

---

## Ideal data

:::: {.columns}

::: {.column width="50%"}

The *ideal* data for 10 people

```{R}
#| echo: FALSE
#| label: "test"
set.seed(3)
ideal_df <- data.frame(
  i = 1:10,
  trt = rep(c(1, 0), each = 5),
  y1i = c(runif(10, 4, 10) %>% round(2)),
  y0i = c(runif(10, 0, 5) %>% round(2))
)
ideal_df
```

:::

::: {.column width="50%"}

Causal effect of treatment.

$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$
for each individual $i$.

:::
::::

---

## Ideal data {data-visibility="uncounted"}

:::: {.columns}

::: {.column width="50%"}

The *ideal* data for 10 people

```{r}
#| label: ideal_data_trt
#| echo: FALSE

ideal_df %>% mutate(effect_i = y1i - y0i)
```

:::

::: {.column width="50%"}

Causal effect of treatment.

$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$

for each individual $i$.

:::

::::

---

## Ideal data {data-visibility="uncounted"}

:::: {.columns}

::: {.column width="50%"}

The *ideal* data for 10 people

```{r}
#| label: ideal_data_trt2
#| echo: FALSE

ideal_df %>% mutate(effect_i = y1i - y0i)

```

:::

::: {.column width="50%"}

Causal effect of treatment.

$$
\begin{align*}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align*}
$$

for each individual $i$.

Define the mean of $\tau_i$ as the [average treatment effect]{.note} (ATE)

$$
\color{#81A1C1}{\overline{\tau} = `r transmute(ideal_df, effect_i = y1i - y0i) %>% unlist %>% mean %>% round(2)`}
$$


:::

::::

. . .

[Notice]{.note} the assignment of treatment is irrelevant in this setting. 

---

## Ideal comparison

$$
\begin{align}
  \tau_i = \color{#81A1C1}{y_{1,i}} &- \color{#B48EAD}{y_{0,i}}
\end{align}
$$

Highlights the fundamental problem of econometrics, much like when a traveler assesses options down two separate roads.

. . .

[The problem]{.note}

- If we observe $\color{#81A1C1}{y_{1,i}}$, then we cannot observe $\color{#B48EAD}{y_{0,i}}$.
- If we observe $\color{#B48EAD}{y_{0,i}}$, then we cannot observe $\color{#81A1C1}{y_{1,i}}$.
- We do not observe the [counterfactual]{.note}.

---

## Counterfactual

> Hypothetical scenario representing the unobserved outcome for an individual or unit if they had experienced the alternative treatment or condition

. . .

The traveler's alternative outcome is forever unknown to them.

---

## Fundamental problem of causal inference

A dataset that we can observe for 10 people looks something like

:::: {.columns}

::: {.column width="50%"}

```{r}
#| label: ideal_data_obs
#| echo: FALSE

obs_df <- ideal_df
obs_df$y0i[1:5] <- NA
obs_df$y1i[6:10] <- NA
obs_df
```

:::

::: {.column width="50%"}

We can't observe $\color{#81A1C1}{y_{1,i}}$ and $\color{#B48EAD}{y_{0,i}}$.

But, we do observe

- $\color{#81A1C1}{y_{1,i}}$ for $i$ in 1, 2, 3, 4, 5
- $\color{#B48EAD}{y_{0,i}}$ for $i$ in 6, 7, 8, 9, 10

:::

::::

. . .

[Q:]{.note} How do we "fill in" the `NA`s and estimate $\overline{\tau}$? [Or.]{.hi .fragment}

. . .

[Q:]{.note} What is a good _counterfactual_ for the missing data?


---

## Estimating causal effects

[Notation:]{.note} $D_i$ is a binary indicator variable such that

- $\color{#81A1C1}{D_i=1}$ _if individual_ $\color{#81A1C1}{i}$ _is treated_.
- $\color{#B48EAD}{D_i=0}$ _if individual_ $\color{#B48EAD}{i}$ _is not treated (control group)_.

. . .

Then, rephrasing the previous slide,

- We only observe $\color{#81A1C1}{y_{1,i}}$ when $\color{#81A1C1}{D_{i}=1}$.
- We only observe $\color{#B48EAD}{y_{0,i}}$ when $\color{#B48EAD}{D_{i}=0}$.

. . .

[Q:]{.note} How can we estimate $\overline{\tau}$ using only $\left(\color{#81A1C1}{y_{1,i}|D_i=1}\right)$ and $\left(\color{#B48EAD}{y_{0,i}|D_i=0}\right)$?

---

## Estimating causal effects

[Q:]{.note} How can we estimate $\overline{\tau}$ using only $\left(\color{#81A1C1}{y_{1,i}|D_i=1}\right)$ and $\left(\color{#B48EAD}{y_{0,i}|D_i=0}\right)$?

. . .

[Idea:]{.note} What if we compare the group of $n$ peoples' means? _I.e._,

$$
\begin{aligned}
  =&\color{#81A1C1}{\mathop{Avg_n}\left( y_i\mid D_i = 1 \right)} - \color{#B48EAD}{\mathop{Avg_n}\left( y_i\mid D_i =0 \right)}\\
  =&\color{#81A1C1}{\mathop{Avg_n}\left( y_{1i}\mid D_i = 1 \right)} - \color{#B48EAD}{\mathop{Avg_n}\left( y_{0i}\mid D_i =0 \right)}
\end{aligned}
$$

. . .

[Q:]{.note} When does a simple difference-in-means provide information on the [causal effect]{.hi} of the treatment?

. . .

[Q:]{.note} Is $\color{#81A1C1}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \color{#B48EAD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$ a *good* estimator for $\overline{\tau}$?


---

## Estimating causal effects

[Assumption:]{.hi} Let $\tau_i = \tau$ for all $i$.

- The treatment effect is equal (constant) across all individuals $i$.

. . .

[Note:]{.note} We defined

$$
\begin{align}
  \tau_i = \tau = \color{#81A1C1}{y_{1,i}} - \color{#B48EAD}{y_{0,i}}
\end{align}
$$

which implies

$$
\begin{align}
   \color{#81A1C1}{y_{1,i}} = \color{#B48EAD}{y_{0,i}} + \tau
\end{align}
$$

---

[Q:]{.note} Is $\color{#81A1C1}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} \color{#434C5E}{-} \color{#B48EAD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$ a *good* estimator for $\tau$?

<br>

. . .

$\quad \color{#ffffff}{\Bigg|}=\color{#81A1C1}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} \color{#434C5E}{-} \color{#B48EAD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$

. . .

$\quad \color{#ffffff}{\Bigg|}=\color{#81A1C1}{\mathop{Avg}\left( y_{1,i}\mid D_i = 1 \right)} \color{#434C5E}{-} \color{#B48EAD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$

. . .

$\quad \color{#ffffff}{\Bigg|}=\color{#B48EAD}{\mathop{Avg}\left( \color{#434C5E}{\tau \: +} \: \color{#B48EAD}{y_{0,i}} \mid D_i = 1 \right)} \color{#434C5E}{-} \color{#B48EAD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$

. . .

$\quad \color{#ffffff}{\Bigg|}= \color{#434C5E}{\tau} \color{#434C5E}{+} \color{#B48EAD}{\mathop{Avg}\left(\color{#B48EAD}{y_{0,i}} \mid D_i = 1 \right)} \color{#434C5E}{-} \color{#B48EAD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$

. . .

$\quad \color{#ffffff}{\Bigg|}= \color{#434C5E}\text{Average causal effect} + \color{#B48EAD}{\text{Selection bias}}$

---

## Estimating causal effects

Our proposed difference-in-means estimator

$$
\color{#434C5E}\text{Average causal effect} + \color{#B48EAD}{\text{Selection bias}}
$$

gives us the sum of

1. $\tau$, the [causal, average treatment effect]{.hi} that we want.
2. [Selection bias:]{.hp} How much treatment and control groups differ, on average.

. . .

[Mastering metrics video on selection bias](https://youtu.be/6YrIDhaUQOE)
