---
name: RCT
---

---

## Selection Bias

[Problem:]{.hi} Selection bias precludes *all else equal* comparisons.

- To make causal statements, we need to shut down the bias term.

. . .

[Potential solution:]{.hi} Conduct an experiment.

- How? [Random assignment of treatment]{.hii}

- Hence the name, [*randomized* control trial]{.note} (RCT).

. . .

Groups will need to be large enough

- Following the LLN, as we increase $n$ of both groups, our randomly assigned treatment estimate is _more likely_ to be representative of the population mean. 


---

## Randomized control trials 

#### [Ex.]{.ex} [Effect of de-worming on attendance]{.hi}

<br>

[Motivation:]{.hi-bb} Intestinal worms are common among children in less-developed countries. The parasitic symptoms disrupt human capital acquisition by keeping children home.

<br>

[Policy Question:]{.hii} Do school-based de-worming interventions provide a cost-effective way to increase school attendance? 


---

## Randomized control trials 

#### [Ex.]{.ex} [Effect of de-worming on attendance]{.hi}

<br>

[Research Question:]{.hi} How much do de-worming interventions increase school attendance?

<br>[Q:]{.note} Could we simply compare average attendance among children with and without access to de-worming medication?

. . .

<br>[A:]{.note} If we're after the causal effect, probably not.

---

## Randomized control trials {data-visibility="uncounted"}

#### [Ex.]{.ex} [Effect of de-worming on attendance]{.hi}

<br>

[Research Question:]{.hi} How much do de-worming interventions increase school attendance?

<br>[Q:]{.note} Why not?

. . .

<br>[A:]{.note} [Selection bias]{.note} [-- Families with access to de-worming medication probably have healthier children for other reasons (_e.g. wealth, access to clean drinking water_).^[_Can't make an *all else equal* comparison. Biased and/or spurious results._]]{.fragment}


---

## Randomized Control Trials 

#### [Ex.]{.ex} [Effect of de-worming on attendance]{.hi}

. . .

Imagine an [RCT]{.note} where we have [two]{.hi} groups of villages:

- [Treatment:]{.hii} Where children get de-worming medication in school.
- [Control:]{.hp} Where children don't get de-worming medication in school _(status quo)_.

. . .

By randomizing, we will, on average, include all kinds of villages in both groups

- poor _vs._ less poor
- access to clean water _vs._ contaminated water
- hospital _vs._ no hospital


---

##

[[54 villages]{.hi}]{.small}

```{r}
#| label: plot1
#| echo: FALSE

xd <- 9
yd <- 6
set.seed(123)
med_df <- expand.grid(x = 1:xd, y = 1:yd) %>%
  mutate(
    q = x + y + 1.25 * rnorm(xd*yd),
    trt = sample(x = c(TRUE, FALSE), size = xd*yd, replace = T, prob = c(0.51, 0.49)),
  ) %>%
  arrange(-y, x) %>%
  mutate(id = 1:(xd*yd) %>% str_pad(2, "left", "0")) %>%
  arrange(y, x)
  
ggplot(data = med_df, aes(x, y)) +
geom_tile(color = "grey40", fill = "grey93", size = 0.2) +
geom_text(aes(label = id), color = "grey45", size = 5, family = "Roboto Mono") +
theme_void() +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_]{.small}

```{r}
#| label: plot2
#| echo: FALSE

ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_1
#| echo: FALSE

set.seed(246)
ggplot(data = med_df, aes(x, y)) +
  geom_tile(aes(fill = q), color = "white", size = 0.2) +
  # geom_text(aes(label = trt)) +
  geom_emoji(data = sample_n(med_df, 1), emoji = "1f48a") +
  scale_fill_viridis(option = "magma", direction = -1) +
  theme_void() +
  theme(legend.position = "none") +
  coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_2
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 2), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_3
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 3), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_4
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 4), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_5
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 5), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```
---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_6
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 6), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_7
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 7), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_8
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 8), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_9
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, 9), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_10
#| echo: FALSE

set.seed(246)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, round(xd * yd / 2, 0)), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_11
#| echo: FALSE

set.seed(248)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, round(xd * yd / 2, 0)), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {data-visibility="uncounted"}

[[54 villages]{.hi} _of varying levels of development_ [plus randomly assigned treatment]{.hii}]{.small}

```{r}
#| label: plot3_12
#| echo: FALSE

set.seed(250)
ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(med_df, round(xd * yd / 2, 0)), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## Randomized Control Trials 

#### [Ex.]{.ex} [Effect of de-worming on attendance]{.hi}

We can estimate the [causal effect]{.hi} of de-worming on school attendance by comparing the average attendance rates in the treatment group (💊) with those in the control group (no 💊).

$$
\begin{align}
  \overline{\text{Attendance}}_\text{Treatment} - \overline{\text{Attendance}}_\text{Control}
\end{align}
$$

. . .


Alternatively, we can use the regression

. . .


$$
\begin{align}
  \text{Attendance}_i = \beta_0 + \beta_1 \text{Treatment}_i + u_i \tag{1}
\end{align}
$$

where $\text{Treatment}_i$ is a binary variable (=1 if village $i$ received the de-worming treatment).

---

## Randomized Control Trials {data-visibility="uncounted"}

#### [Ex.]{.ex} [Effect of de-worming on attendance]{.hi}

$$
\begin{align}
  \text{Attendance}_i = \beta_0 + \beta_1 \text{Treatment}_i + u_i \tag{1}
\end{align}
$$

where $\text{Treatment}_i$ is a binary variable (=1 if village $i$ received the de-worming treatment).


[Q:]{.note} Should trust the results of Eq. $(1)$? [Why?]{.fragment}

. . .

<br>[A:]{.note} On average, [randomly assigning treatment should balance]{.hi} treatment and control across the other dimensions that affect school attendance.


---

[Randomization can go wrong!]{.hi}

```{r}
#| label: plot3_bad
#| echo: FALSE

ggplot(data = med_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = filter(med_df, x < xd/3 | y <= yd/3 | (x == xd & y == yd)), emoji = "1f48a") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```