---
name: math-review
---


---

## Notation

Data on a variable $X$ are a sequence of $n$ observations, indexed by $i$: $$\{x_i: 1, \dots, n \}.$$

. . .

[Ex.]{.ex} $n = 5$

:::: {.columns}

::: {.column width="35%"}
```{r}
#| echo = FALSE
# Sample size
n <- 5

set.seed(94665)

# Generate datata
table_df <- data.frame(
  i = 1:n,
  x = sample(1:10, n) # draw integers
)

colnames(table_df) <- c("\\(i\\)", "\\(x_i\\)")

table_df %>%
  kbl() %>%
  kable_minimal(full_width = FALSE, position = "left")
```
:::

::: {.column width="65%"}
<br>
<br>

- $i$ indicates the row number.

- $n$ is the number of rows.

- $x_i$ is the value of $X$ for row $i$.
:::

::::

---

## Summation

The [summation operator]{.hii} adds a sequence of numbers over an index:

$$
\sum_{i=1}^{n} x_i \equiv x_1 + x_2 + \dots + x_n.
$$

> The sum of $x_i$ from 1 to $n$.

. . .

:::: {.columns}

::: {.column width="30%"}


```{r, echo=FALSE, include=TRUE, results="asis"}
df <- data.frame(
  i = 1:4,
  x = c(7, 4, 10, 3)
)

colnames(df) <- c("\\(i\\)", "\\(x_i\\)")

df %>%
  kbl() %>%
  kable_minimal(full_width = FALSE, position = "left")
```
:::

::: {.column width="70%"}
$$
\begin{aligned}
 \sum_{i=1}^{4} x_i = 7 + 4 + 10 + 3 = 23 \\
  \frac{1}{n} \sum_{i=1}^n x_i  \rightarrow \frac{1}{4} \sum_{i=1}^4 x_i &= \  6  
\end{aligned}
$$

:::
::::

---

## Summation {data-visibility="uncounted"}

The [summation operator]{.hii} adds a sequence of numbers over an index:

$$
\sum_{i=1}^{n} x_i \equiv x_1 + x_2 + \dots + x_n.
$$

> The sum of $x_i$ from 1 to $n$.

:::: {.columns}

::: {.column width="30%"}

```{r, echo=FALSE, include=TRUE, results="asis"}
df <- data.frame(
  i = 1:4,
  x = c(7, 4, 10, 3)
)

colnames(df) <- c("\\(i\\)", "\\(x_i\\)")

df %>%
  kbl() %>%
  kable_minimal(full_width = FALSE, position = "left")
```
:::

::: {.column width="70%"}
$$
\begin{aligned}
 \sum_{i=1}^{4} x_i = 7 + 4 + 10 + 3 = 23 \\
 {\scriptsize \text{sample average}} \Bigg \{ \frac{1}{n} \sum_{i=1}^n x_i  \rightarrow \frac{1}{4} \sum_{i=1}^4 x_i &= \  6  
\end{aligned}
$$

:::
::::


---

## Summation: Rule 01

For any constant $c$, 

$$
\sum_{i=1}^{n} c = nc.
$$

. . .

:::: {.columns}
::: {.column width="30%"}

```{r, echo=FALSE, include=TRUE, results="asis"}
df <- data.frame(
  i = 1:4,
  c = c(2, 2, 2, 2)
)

colnames(df) <- c("\\(i\\)", "\\(c\\)")

df %>%
  kbl() %>%
  kable_minimal(full_width = FALSE, position = "left")
```
:::

::: {.column width="70%"}

$$
\begin{aligned}
 \sum_{i=1}^{4} 2 &= 4 \times 2 \\
                  &= 8
\end{aligned}
$$

:::
::::


---

## Summation: Rule 02

For any constant $c$, $$\sum_{i=1}^{n} cx_i = c \sum_{i=1}^{n} x_i.$$

. . .

:::: {.columns}
::: {.column width="30%"}

<br>

```{r, echo=FALSE, include=TRUE, results="asis"}
df <- data.frame(
  i = 1:3,
  c = c(2, 2, 2),
  x = c(7, 4, 10) 
)

colnames(df) <- c("\\(i\\)", "\\(c\\)", "\\(x_i\\)")

df %>%
  kbl() %>%
  kable_minimal(full_width = FALSE, position = "left")
```
:::

::: {.column width="70%"}
\begin{align}
 \sum_{i=1}^{3} 2x_i &= 2\times7 + 2\times4 + 2 \times10\\
               &= 14 + 8 + 20 = 42 \\

 2 \sum_{i=1}^{3} x_i &= 2(7 + 4 + 10) = 42
\end{align}
:::
::::


---

## Summation: Rule 03

If $\{(x_i, y_i): 1, \dots, n \}$ is a set of $n$ pairs, and $a$ and $b$ are constants, then $$\sum_{i=1}^{n} (ax_i + by_i) = a \sum_{i=1}^{n} x_i + b \sum_{i=1}^{n} y_i.$$

. . .

:::: {.columns}
::: {.column width="30%"}

<br>

```{r, echo=FALSE, include=TRUE, results="asis"}
df <- data.frame(
  i = 1:2,
  a = c(2, 2),
  x = c(7, 4),
  b = c(1, 1),
  y = c(4, 2)
)

colnames(df) <- c("\\(i\\)", "\\(a\\)", "\\(x_i\\)", "\\(b\\)", "\\(y_i\\)")

df %>%
  kbl() %>%
  kable_minimal(full_width = FALSE, position = "left")
```
:::

::: {.column width="70%"}
\begin{align}
 \sum_{i=1}^{2} (2x_i + y_i) &= 18 + 10 = 28 \\
 2 \sum_{i=1}^{2} x_i + \sum_{i=1}^{2} y_i &= 2 \times 11 + 6 = 28
\end{align}
:::
::::


---

## Summation: Caution 01

The [sum of the ratios]{.hii} is not the [ratio of the sums:]{.hp} 
$$
{\color{#81A1C1}{\sum_{i=1}^{n} x_i / y_i}} \neq \color{#B48EAD}{\left(\sum_{i=1}^{n} x_i \right) \Bigg/ \left(\sum_{i=1}^{n} y_i \right)}
$$

. . .

[Ex.]{.ex}

If $n = 2$, then $\frac{x_1}{y_1} + \frac{x_2}{y_2} \neq \frac{x_1 + x_2}{y_1 + y_2}$.

---

## Summation: Caution 02

The [sum of squares]{.hii} is not the [square of the sums:]{.hp}
$$
\color{#81A1C1}{\sum_{i=1}^{n} x_i^2} \neq \color{#B48EAD}{\left(\sum_{i=1}^{n} x_i \right)^2}.
$$

. . .

[Ex.]{.ex}

If $n = 2$, then $x_1^2 + x_2^2 \neq (x_1 + x_2)^2 = x_1^2 + 2x_1x_2 + x_2^2$.

---

## Cartesian coordinate system

Cartesian plane: 2-D plane defined by two perpendicular number lines:

- x-axis (_horizontal_)
- y-axis (_vertical_) 
  
. . .

Using these axes, any point in the plane is described using an ordered pair of numbers $(x,y)$

---

## Cartesian coordinate system

A particular line on this plane takes the form
$$
\begin{aligned}
y= a+ bx
\end{aligned}
$$
where $a$ is known as the intercept and $b$ is the slope.

Any incremental unit increase in $x$ results in $y$ increasing by $b$.

---

## [Ex.]{.ex}

```{R}
#| label: "spurious"
#| fig-align: center
#| echo: FALSE
#| fig-width: 16
#| fig-height: 10

cheese <- c(29.8,30.1,30.5,30.6,31.3,31.7,32.6,33.1,32.7,32.8)
bed_ded <- c(327,456,509,497,596,573,661,741,809,717)

df <- data.frame(cheese=cheese, 
                 bed_ded=bed_ded)

ggplot(df, aes(cheese, bed_ded)) +
  geom_point(size = 4, color = "darkslategray") +
  geom_smooth(method = "lm", se = FALSE, color = hii) +
  xlab("Per Capita Cheese Consumption (lbs.)") +
  ylab("Death by bedsheet tangling") + 
  mytheme +
  theme(
    axis.text = element_text(size = 18)
  )
```