---
title: "Simple linear regression II"
subtitle: "EC 320, Set 05"
author: "Andrew Dickinson"
date: last-modified
date-format: "Spring YYYY"
format: 
  revealjs:
    theme: [default, ../styles.scss]
    monobackgroundcolor: #F5F5F5
    slide-number: true
    footer: "EC320, Set 05 | Simple linear regression (again)"
    preview-links: auto
    code-fold: FALSE
    # code-copy: TRUE
    highlight-style: a11y-light
    cache: TRUE
    html-math-method: mathjax 
title-slide-attributes: 
  data-background-position: left
---

::: {.content-hidden}
$$
{{< include ../_macros.tex >}}
$$
:::

## Housekeeping

[PS03:]{.hi}

- Due _(next)_ Tuesday at 11:59p
- Covers remaining a bunch of OLS stuff up to today

. . .

[LA04:]{.hi}

- Due Friday at 5:00p

. . .

[Reading:]{.hi} _(up to this point)_

ItE: R, 1, [2]{.hi}
MM: 1, [2]{.hi}

# _Prologue_ {.inverse .note}

```{r}
pacman::p_load(hrbrthemes, fastverse, tidyverse,
               magrittr, wooldridge, here, kableExtra,
               ggdag, nord, latex2exp, dagitty, viridis,
               plotly, ggforce, latex2exp, parallel, broom)


hi = nord_palettes$polarnight[3]
hii = nord_palettes$frost[3] 
hp = nord_palettes$aurora[5]
higreen = nord_palettes$aurora[4]
hiyellow = nord_palettes$aurora[3]
hiorange = nord_palettes$aurora[2]
hired = nord_palettes$aurora[1]
higrey = nord_palettes$snowstorm[1]

mytheme = theme_ipsum(base_family = "Fira Sans Book", base_size = 20) +
 theme(panel.grid.minor.x = element_blank(),
       axis.title.x = element_text(size = 18),
       axis.title.y = element_text(size = 18))

mytheme_s = mytheme + 
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.line = element_line(color = hi))


# simulated earnings data
df_earn <- tibble(x = sample(8:20, 100, replace=T),
             y = -14 + 2.5*x + rnorm(100, 0, 5)) %>% 
  filter(y > 5)
lm_earn <- lm(y ~ x, data = df_earn)
b0_earn <- lm_earn$coefficients[1]
b1_earn <- lm_earn$coefficients[2]
# data for R^2 discussion
df1 <- tibble(x = rnorm(1000, 5, 5),
             y = 100 - 1*x + rnorm(1000, 0, 3))
df2 <- tibble(x = rnorm(1000, 5, 5),
             y = 100 - 1*x + rnorm(1000, 0, 20))
r2_1 <- summary(lm(y~x, df1))$r.squared
r2_2 <- summary(lm(y~x, df2))$r.squared
# campus crime data
data <- get(data(campus)) %>% 
  mutate(crime = round(crime/enroll*1000, 2),
         police = round(police/enroll*1000, 2)) %>% 
  filter(police < 10) %>% # remove outlier
  select(crime, police)
data2 <- data %>% 
  mutate(crime = crime/1000)
lm0 <- lm(crime ~ police, data = data)
lm1 <- lm(crime ~ police, data = data2)

```

{{< include _041-prologue.qmd >}}

# _OLS properties_ {.inverse .note}

{{< include _042-properties.qmd >}}

# _Goodness of fit_ {.inverse .note}

{{< include _043-goodness.qmd >}}

# _Units of measurement_ {.inverse .note}

{{< include _044-units.qmd >}}

# [Ex.]{.ex} [Returns to schooling]{.hi-white} {.inverse}

{{< include _045-ex.qmd >}}

# _Classical assumptions_ {.inverse .note}

{{< include _046-assumptions.qmd >}}

# _Gauss Markov_ {.inverse .note}

{{< include _047-gaussmarkov.qmd >}}

# _Population vs. sample, revisited_ {.inverse .note}

{{< include _048-sims.qmd >}}

## 

::: {.vertical-center}

We will further this discussion next time to conduct inference.

:::

# Fin {.inverse}

```{r}
#| include: false
#| echo: false
#| eval: false
library(renderthis)

renderthis::to_pdf(from = "./slides/004-ols-ii/050-main.html", to = "./slides/004-ols-ii/050-main.pdf")
```