---
name: goodness
---

---

## Goodness of Fit

[Regression 1]{.hi-orange} *vs.* [Regression 2]{.hii}

- Same slope.
- Same intercept. 

[Q:]{.note} Which fitted regression line *"explains"*^[_Explains_ [=]{.mono} _fits_.] the data better?


:::: {.columns}

::: {.column width="50"}
```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

ggplot() +
  xlim(-15,25) + ylim(0, 160) +
  geom_point(data = df1, aes(x, y), color = hi, alpha = 0.5, size = 4) +
  stat_smooth(data = df1, aes(x, y), method = "lm", se = FALSE, color = hiorange, size = 5) +
  theme_void()
```
:::

::: {.column width="50"}
```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

ggplot() +
  xlim(-15,25) + ylim(0, 160) +
  geom_point(data = df2, aes(x, y), color = hi, alpha = 0.5, size = 4) +
  stat_smooth(data = df2, aes(x, y), method = "lm", se = FALSE, color = hii, size = 5) +
  theme_void()
```
:::

::::

---

## Goodness of Fit {data-visibility="uncounted"}

[Regression 1]{.hi-orange} *vs.* [Regression 2]{.hii}

The [coefficient of determination]{.note}, $R^2$, is the fraction of the variation in $Y_i$ *"explained"* by $X_i$.

- $R^2 = 1 \implies X_i$ explains _all_ of the variation in $Y_i$.
- $R^2 = 0 \implies X_i$ explains _none_ of the variation in $Y_i$.

:::: {.columns}

::: {.column width="50"}
```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center
#| fig-cap: !expr 'paste("$R^2$ =", round(r2_1,2))'
#| fig-cap-location: bottom

ggplot() +
  xlim(-15,25) + ylim(0, 160) +
  geom_point(data = df1, aes(x, y), color = hi, alpha = 0.5, size = 4) +
  stat_smooth(data = df1, aes(x, y), method = "lm", se = FALSE, color = hiorange, size = 5) +
  theme_void()
```
:::

::: {.column width="50"}
```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center
#| fig-cap: !expr 'paste("$R^2$ =", round(r2_2,2))'

ggplot() +
  xlim(-15,25) + ylim(0, 160) +
  geom_point(data = df2, aes(x, y), color = hi, alpha = 0.5, size = 4) +
  stat_smooth(data = df2, aes(x, y), method = "lm", se = FALSE, color = hii, size = 5) +
  theme_void()
```
:::

::::

---

## 

```{r}
#| echo: false
#| fig.height: 7
#| fig.align: center


# Colors (order: x1, y)
venn_colors <- c(hp, hii)

# Locations of circles
venn_df <- tibble(
  x  = c( 0.0,   -0.5),
  y  = c( 0.0,   -2.5),
  r  = c( 1.9,    1.9),
  l  = c( "Y", "X"),
  xl = c( 0.0,   -0.5),
  yl = c( 0.0,   -2.5)
)
# Venn
ggplot(data = venn_df, aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
geom_text(aes(x = xl, y = yl, label = l), size = 12, family = "Fira Sans", parse = T) +
xlim(-2.5, 2) +
ylim(-4.5, 2) +
coord_equal()
```

---

## {data-visibility="uncounted"}

```{r}
#| echo: false
#| fig.height: 7
#| fig.align: center

# Locations of circles
venn_df <- tibble(
  x  = c( 0.0,   -3),
  y  = c( 0.0,   -2.75),
  r  = c( 1.9,    1.9),
  l  = c( "Y", "X"),
  xl = c( 0.0,   -3),
  yl = c( 0.0,   -2.75)
)
# Venn
ggplot(data = venn_df, aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
geom_text(aes(x = xl, y = yl, label = l), size = 12, family = "Fira Sans", parse = T) +
xlim(-5, 2) +
ylim(-5, 2) +
coord_equal()
```

---

## {data-visibility="uncounted"}

```{r}
#| echo: false
#| fig.height: 7
#| fig.align: center

# Locations of circles
venn_df <- tibble(
  x  = c( 0.0,   -0.5),
  y  = c( 0.0,   -1.5),
  r  = c( 1.9,    1.9),
  l  = c( "Y", "X"),
  xl = c( 0.0,   -0.5),
  yl = c( 1.0,   -2.5)
)
# Venn
ggplot(data = venn_df, aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
geom_text(aes(x = xl, y = yl, label = l), size = 12, family = "Fira Sans", parse = T) +
xlim(-2.5, 2) +
ylim(-5, 2) +
coord_equal()
```

---

## Explained and unexplained variation

Residuals remind us that there are parts of $Y_i$ we can't explain.

$$
Y_i = \hat{Y_i} + \hat{u}_i
$$

- Sum the above, divide by $n$, and use the fact that OLS residuals sum to zero to get:

$$
\bar{\hat{u}} = 0 \implies \bar{Y} = \bar{\hat{Y}}
$$

---

## Explained and unexplained variation

[Total Sum of Squares]{.note} ([TSS]{.hi-red}) measures variation in $Y_i$:

$$
{\color{#BF616A} \text{TSS}} \equiv \sum_{i=1}^n (Y_i - \bar{Y})^2
$$

- [TSS]{.hi-red} can be decomposed into explained and unexplained variation.

. . .

:::: {.columns}

::: {.column width="50%"}
[Explained Sum of Squares]{.note} ([ESS]{.hi-yellow}) measures the variation in $\hat{Y_i}$:

$$
{\color{#EBCB8B} \text{ESS}} \equiv \sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2
$$
:::

::: {.column width="50%"}
[Residual Sum of Squares]{.note} ([RSS]{.hi-orange}) measures the variation in $\hat{u}_i$:

$$
{\color{#D08770} \text{RSS}} \equiv \sum_{i=1}^n \hat{u}_i^2
$$
:::

::::

---

## {data-visibility="uncounted"}

::: {.vertical-center}
[Goal:]{.note} Show that ${\color{#BF616A} \text{TSS}} = {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}}$.
:::

---

## {data-visibility="uncounted"}

[Goal:]{.note} Show that ${\color{#BF616A} \text{TSS}} = {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}}$.

[_Step 1:_]{.note} Plug $Y_i = \hat{Y_i} + \hat{u}_i$ into ${\color{#BF616A} \text{TSS}}$.

$$
\begin{align*}
{\color{#BF616A} \text{TSS}} &= \sum_{i=1}^n (Y_i - \bar{Y})^2 \\
                             &= \sum_{i=1}^n ([\hat{Y_i} + \hat{u}_i] - [\bar{\hat{Y}} + \bar{\hat{u}}])^2
\end{align*}
$$

---

## {data-visibility="uncounted"}

[Goal:]{.note} Show that ${\color{#BF616A} \text{TSS}} = {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}}$.

[_Step 2:_]{.note} Recall that $\bar{\hat{u}} = 0$ & $\bar{Y} = \bar{\hat{Y}}$.

$$
\begin{align*}
{\color{#BF616A} \text{TSS}} &= \sum_{i=1}^n \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right)^2 \\
                             &= \sum_{i=1}^n \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right) \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right) \\
                             &= \sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2 + \sum_{i=1}^n \hat{u}_i^2 + 2 \sum_{i=1}^n \left( (\hat{Y_i} - \bar{Y})\hat{u}_i \right)
\end{align*}
$$

<br>

---

## {data-visibility="uncounted"}

[Goal:]{.note} Show that ${\color{#BF616A} \text{TSS}} = {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}}$.

[_Step 2:_]{.note} Recall that $\bar{\hat{u}} = 0$ & $\bar{Y} = \bar{\hat{Y}}$.

$$
\begin{align*}
{\color{#BF616A} \text{TSS}} &= \sum_{i=1}^n \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right)^2 \\
                             &= \sum_{i=1}^n \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right) \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right) \\
                             &= {\color{#EBCB8B}{\sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2}} + {\color{#D08770}{\sum_{i=1}^n \hat{u}_i^2}} + 2 \sum_{i=1}^n \left( (\hat{Y_i} - \bar{Y})\hat{u}_i \right)
\end{align*}
$$

[_Step 3:_]{.note} Notice [ESS]{.hi-yellow} and [RSS]{.hi-orange}.

---

## {data-visibility="uncounted"}

[Goal:]{.note} Show that ${\color{#BF616A} \text{TSS}} = {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}}$.

[_Step 2:_]{.note} Recall that $\bar{\hat{u}} = 0$ & $\bar{Y} = \bar{\hat{Y}}$.

$$
\begin{align*}
{\color{#BF616A} \text{TSS}} &= \sum_{i=1}^n \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right)^2 \\
                             &= \sum_{i=1}^n \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right) \left( [\hat{Y_i} - \bar{Y}] + \hat{u}_i \right) \\
                             &= {\color{#EBCB8B}{\sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2}} + {\color{#D08770}{\sum_{i=1}^n \hat{u}_i^2}} + 2 \sum_{i=1}^n \left( (\hat{Y_i} - \bar{Y})\hat{u}_i \right) \\
                             &= {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}} + 2 \sum_{i=1}^n \left( (\hat{Y_i} - \bar{Y})\hat{u}_i \right)
\end{align*}
$$

---

## {data-visibility="uncounted"}

[Goal:]{.note} Show that ${\color{#BF616A} \text{TSS}} = {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}}$.

[_Step 4:_]{.note} Simplify.

$$
\begin{align*}
{\color{#BF616A} \text{TSS}} &= {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}} + 2 \sum_{i=1}^n \left( (\hat{Y_i} - \bar{Y})\hat{u}_i \right) \\
                             &= {\color{#EBCB8B} \text{ESS}} + {\color{#D08770} \text{RSS}} + 2 \sum_{i=1}^n \hat{Y_i}\hat{u}_i - 2 \bar{Y}\sum_{i=1}^n \hat{u}_i
\end{align*}
$$

. . .

[_Step 5:_]{.note} Shut down the last two terms. Notice that

$$
\begin{align*}
2 \sum_{i=1}^n \hat{Y_i}\hat{u}_i - 2 \bar{Y}\sum_{i=1}^n \hat{u}_i = 0
\end{align*}
$$

_Which you will prove to be true in [PS03]{.hi}_

---

##

::: {.vertical-center}
Some visual intuition makes all the math seem a lot simpler

<br>

[Let's regress `mpg` on `wt` using a new dataset, `mtcars`]{.fragment}
:::

---

:::: {.columns}

::: {.column width="50%"}
$$
{\color{#ffffff} \text{TSS} \equiv \sum_{i=1}^n (Y_i - \bar{Y})^2}
$$
:::

::: {.column width="50%"}
$$
{\color{#ffffff} \text{TSS} \equiv \sum_{i=1}^n (Y_i - \bar{Y})^2}
$$
:::

::::

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

mtcars %>% 
  ggplot(aes(x = wt, y = mpg)) +
    geom_point(color = hi) +
    mytheme_s +
    coord_fixed() +
    coord_cartesian(xlim = c(1,6), ylim = c(0,35)) +
    xlab("Weight (thousands)") + 
    ylab("MPG")
```

---

:::: {.columns}

::: {.column width="30%"}
$$
{\color{#A3BE8C} \overline{\text{MPG}_i}} = 20.09
$$
:::

::: {.column width="70%"}
$$
{\color{#ffffff} \text{TSS} \equiv \sum_{i=1}^n (Y_i - \bar{Y})^2}
$$
:::

::::

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center


lm0 <- lm(mpg ~ wt, data = mtcars)

# Define a function
y_hat <- function(x, b0, b1) {b0 + b1 * x}
# Define line's parameters
b0 <- lm0$coefficients[1]
b1 <- lm0$coefficients[2]

y_bar = mtcars$mpg %>% mean()

mtcars %>% 
  ggplot(aes(x = wt, y = mpg)) +
    geom_hline(yintercept = y_bar, color = higreen, size = 1) +
    geom_point(color = hi) +
    coord_cartesian(xlim = c(1,6), ylim = c(0,35)) +
    scale_size(range = c(0.001,2)) +
    # scale_color_continuous() +
    mytheme_s +
    xlab("Weight (thousands)") + 
    ylab("MPG") +
    theme(
      legend.position = 'none'
    )
```

---

:::: {.columns}

::: {.column width="30%"}
$$
{\color{#BF616A} \text{TSS}} \equiv \sum_{i=1}^n (Y_i - \bar{Y})^2
$$
:::

::: {.column width="70%"}
$$
{\color{#ffffff} \overline{\text{Crime}_i} = 21.05}
$$
:::

::::

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

mtcars %>% 
  ggplot(aes(x = wt, y = mpg)) +
    geom_hline(yintercept = y_bar, color = higreen, size = 1) +
    geom_point(color = hi) +
    # geom_abline(intercept = b0, slope = b1, color = hp, size = 1.25) +
    geom_segment(aes(x = wt, xend = wt, y = mpg, yend = y_bar), color = hired, alpha = 1, size = 1) +
    # coord_fixed() +
    coord_cartesian(xlim = c(1,6), ylim = c(0,35)) +
    scale_size(range = c(0.001,2)) +
    # scale_color_continuous() +
    mytheme_s +
    xlab("Weight (thousands)") + 
    ylab("MPG") +
    theme(
      legend.position = 'none'
    )
```

---

:::: {.columns}

::: {.column width="40%"}
$$
{\color{#B48EAD} \widehat{\text{Crime}_i}} = 18.41 + 1.76 \cdot \text{Police}_i 
$$
:::

::: {.column width="60%"}
$$
{\color{#ffffff} \text{ESS} \equiv \sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2}
$$
:::

::::


```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

mtcars %>% 
  ggplot(aes(x = wt, y = mpg)) +
    geom_point(color = hi) +
    geom_hline(yintercept = y_bar, color = higreen, size = 1) +
    geom_abline(intercept = b0, slope = b1, color = hp, size = 1.25) +
    # coord_fixed() +
    coord_cartesian(xlim = c(1,6), ylim = c(0,35)) +
    scale_size(range = c(0.001,2)) +
    # scale_color_continuous() +
    mytheme_s +
    xlab("Weight (thousands)") + 
    ylab("MPG") +
    theme(
      legend.position = 'none'
    )
```

---

:::: {.columns}

::: {.column width="40%"}
$$
{\color{#EBCB8B} \text{ESS}} \equiv \sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2
$$
:::

::: {.column width="60%"}
$$
{\color{#ffffff} \widehat{\text{Crime}_i} = 18.41 + 1.76 \cdot \text{Police}_i}
$$
:::

::::


```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

mtcars %>% 
  ggplot(aes(x = wt, y = mpg)) +
    geom_point(color = hi) +
    geom_hline(yintercept = y_bar, color = higreen, size = 1) +
    geom_abline(intercept = b0, slope = b1, color = hp, size = 1.25) +
    geom_segment(aes(x = wt, xend = wt, y = y_bar, yend = y_hat(wt, b0, b1)), color = hiyellow, alpha = 1, size = 1) +
    # coord_fixed() +
    coord_cartesian(xlim = c(1,6), ylim = c(0,35)) +
    scale_size(range = c(0.001,2)) +
    # scale_color_continuous() +
    mytheme_s +
    xlab("Weight (thousands)") + 
    ylab("MPG") +
    theme(
      legend.position = 'none'
    )
```


---

:::: {.columns}

::: {.column width="40%"}
$$
{\color{#D08770} \text{RSS}} \equiv \sum_{i=1}^n \hat{u}_i^2
$$
:::

::: {.column width="60%"}
$$
{\color{#ffffff} \text{ESS} \equiv \sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2}
$$
:::

::::


```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

mtcars %>% 
  ggplot(aes(x = wt, y = mpg)) +
    geom_point(color = hi) +
    geom_hline(yintercept = y_bar, color = higreen, size = 1) +
    geom_abline(intercept = b0, slope = b1, color = hp, size = 1.25) +
    geom_segment(aes(x = wt, xend = wt, y = mpg, yend = y_hat(wt, b0, b1)), color = hiorange, alpha = 1, size = 1) +
    # coord_fixed() +
    coord_cartesian(xlim = c(1,6), ylim = c(0,35)) +
    scale_size(range = c(0.001,2)) +
    # scale_color_continuous() +
    mytheme_s +
    xlab("Weight (thousands)") + 
    ylab("MPG") +
    theme(
      legend.position = 'none'
    )
```

---

:::: {.columns}

::: {.column width="33%"}
$$
{\color{#BF616A} \text{TSS}} \equiv \sum_{i=1}^n (Y_i - \bar{Y})^2
$$
:::

::: {.column width="33%"}
$$
{\color{#EBCB8B} \text{ESS}} \equiv \sum_{i=1}^n (\hat{Y_i} - \bar{Y})^2
$$
:::

::: {.column width="33%"}
$$
{\color{#D08770} \text{RSS}} \equiv \sum_{i=1}^n \hat{u}_i^2
$$
:::

::::

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

mtcars %>% 
  ggplot(aes(x = wt, y = mpg)) +
    geom_hline(yintercept = y_bar, color = higreen, size = 1) +
    geom_point(color = hi) +
    geom_abline(intercept = b0, slope = b1, color = hp, size = 1.25) +
    geom_segment(aes(x = wt, xend = wt, y = mpg, yend = y_bar), color = hired, position = position_dodge(5), alpha = 1, size = 1) +
    geom_segment(aes(x = wt-0.02, xend = wt-0.02, y = mpg, yend = y_hat(wt, b0, b1)), color = hiorange, alpha = 1, size = 1) +
    geom_segment(aes(x = wt+0.02, xend = wt+0.02, y = y_bar, yend = y_hat(wt, b0, b1)), color = hiyellow, alpha = 1, size = 1) +
    # coord_fixed() +
    coord_cartesian(xlim = c(1,6), ylim = c(0,35)) +
    scale_size(range = c(0.001,2)) +
    # scale_color_continuous() +
    mytheme_s +
    xlab("Weight (thousands)") + 
    ylab("MPG") +
    theme(
      legend.position = 'none'
    )
```


---

## Goodness of fit

What percentage of the variation in our $Y_i$ is *apparently* explained by our model? The $R^2$ term represents this percentage.

Total variation is represented by [TSS]{.hi-red} and our model is capturing the 'explained' sum of squares, [ESS]{.hi-yellow}.

Taking a simple ratio reveals how much variation our model explains. 

- $R^2 = \frac{{\color{#EBCB8B} \text{ESS}}}{{\color{#BF616A} \text{TSS}}}$ varies between 0 and 1

- $R^2 = 1 - \frac{{\color{#D08770} \text{RSS}}}{{\color{#BF616A} \text{TSS}}}$, 100% less the unexplained variation 

. . .

$R^2$ is related to the correlation between the actual values of $Y$ and the fitted values of $Y$. Can show that $R^2 = (r_{Y, \hat{Y}})^2$.

---

## Goodness of fit

_So what?_ [In the social sciences, low $R^2$ values are common.]{.fragment}

. . .

Low $R^2$ doesn't mean that an estimated regression is useless.

- In a randomized control trial, $R^2$ is usually less than 0.1

. . .

High $R^2$ doesn't necessarily mean you have a *"good"* regression.

- Worries about selection bias and omitted variables still apply
- Some 'powerfully high' $R^2$ values are the result of simple accounting exercises, and tell us nothing about causality

