---
name: ex
---

---

## Returns to Schooling

```{r}
#| echo: false


# Neumark data
wage <- get(data(wage2))
lm_e <- lm(lwage ~ educ, data = wage)
b0 <- lm_e$coefficients[1]
b1 <- lm_e$coefficients[2]
r_2 <- summary(lm(lwage ~ educ, data = wage))$r.squared
```

[Policy Question:]{.note} _How much should the state subsidize higher education?_

- Could higher education subsidies increase future tax revenue?
- Could targeted subsidies reduce income and racial inequality?
- Are there positive externalities associated with higher education? 

. . .

<br>

[_Empirical Question:_]{.note} _What is the return to an additional year of education?_^[Useful for learning about the econometric assumptions that allow causal interpretation.]

---

## Returns to Schooling

[Step 1:]{.note} Write down the population model.

$$
\log(\text{Earnings}_i) = \beta_1 + \beta_2\text{Education}_i + u_i
$$

. . .

[Step 2:]{.note} Find data.

- *Source:* [Blackburn and Neumark (1992)](https://econpapers.repec.org/article/oupqjecon/v_3a107_3ay_3a1992_3ai_3a4_3ap_3a1421-1436..htm).

. . .

[Step 3:]{.note} Run a regression using OLS.

$$
\log(\hat{\text{Earnings}_i}) = \hat{\beta}_1 + \hat{\beta}_2\text{Education}_i
$$

# [Aside:]{.note} Logarithms

---

## [Aside:]{.note} Logarithms

[Logarithms]{.note} are functions are convenient for numbers that span several orders of magnitude

<br>

. . .

The [*natural logarithm*]{.note}

> Inverse function to the exponential function with base $e$ defined as:

$$
\log{x} \quad \textit{or} \quad \ln{x}
$$

. . .

Simplifies interpretation and stabilize relationships between variables.

---

## [Aside:]{.note} Logarithms

Our example model: $$\log(\text{Earnings}_i) = \beta_1 + \beta_2\text{Education}_i + u_i$$

[Natural logarithm]{.note} is applied to the dependent variable (Earnings).

. . .

<br>

Estimates how the *percentage change* in earnings. [$\beta_2$ represents the percentage change in earnings for a one-unit increase in education.]{.fragment}

. . .

# [_Return to our_]{.note} [example]{.ex} {data-visibility="uncounted"}

---

## Returns to Schooling

$\log(\hat{\text{Earnings}_i})$ $=$ [`r round(b0, 2)`]{.hp} $+$ [`r round(b1, 2)`]{.hp} $\times$ $\text{Education}_i$.

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

wage %>% 
  ggplot() +
  geom_point(aes(x = educ, y = lwage), color = hi) +
  geom_abline(intercept = b0, slope = b1, color = hp, size = 1) +
  mytheme_s + xlab("Years of Education") + ylab("log(Monthly Earnings)")
```

---

## Returns to Schooling

Additional year of school associated with a [`r round(b1, 2)*100`%]{.hp} increase in earnings. 

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

wage %>% 
  ggplot() +
  geom_point(aes(x = educ, y = lwage), color = hi) +
  geom_abline(intercept = b0, slope = b1, color = hp, size = 1) +
  mytheme_s + xlab("Years of Education") + ylab("log(Monthly Earnings)")
```

---

## Returns to Schooling

$R^2$ $=$ [`r round(r_2, 3)`]{.hp}. 

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

wage %>% 
  ggplot() +
  geom_point(aes(x = educ, y = lwage), color = hi) +
  geom_abline(intercept = b0, slope = b1, color = hp, size = 1) +
  mytheme_s + xlab("Years of Education") + ylab("log(Monthly Earnings)")
```

---

## Returns to Schooling

Education explains [`r round(r_2, 3)*100`%]{.hp} of the variation in wages. 

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

wage %>% 
  ggplot() +
  geom_point(aes(x = educ, y = lwage), color = hi) +
  geom_abline(intercept = b0, slope = b1, color = hp, size = 1) +
  mytheme_s + xlab("Years of Education") + ylab("log(Monthly Earnings)")
```

---

## Returns to Schooling

What must we [_assume_]{.hi} to interpret $\hat{\beta}_2$ $=$ [`r round(b1, 2)`]{.hp} as the return to schooling?

```{r}
#| echo: false
#| fig.height: 5.75
#| fig.align: center

wage %>% 
  ggplot() +
  geom_point(aes(x = educ, y = lwage), color = hi) +
  geom_abline(intercept = b0, slope = b1, color = hp, size = 1) +
  mytheme_s + xlab("Years of Education") + ylab("log(Monthly Earnings)")
```

