---
title: "**Introduction to R**"
author: | 
        | **Modifications:** Giorgi Nikolaishvili
        | **Original Authors:** Kyle Raze, Youssef Ait Benasser, & Saurabh Gupta
        | EC 320: Introduction to Econometrics
        | University of Oregon
date: "Spring 2022"
output: 
  html_document:
    theme: flatly  
    includes:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, cache = T)
```

## Labs vs. Lectures

**Lectures:** learning new econometric concepts, math, research design, etc. 

**Labs:** gathering tools to apply the above concepts to real data.

## A Little Bit About Me

Full name: Giorgi Nikolaishvili -- you can either refer to me as Gio ("G" pronounced as in "great" or "Google"), or George.

Tbilisi, Georgia -> Washington, D.C. -> Tbilisi, Georgia -> Kyiv, Ukraine -> Boston, MA -> Eugene, OR.

Like Emmett, I'm also a 3rd-year Economics PhD Candidate.

Research areas: macroeconomics & time series econometrics. 

I study how the banking and financial sectors evolve over time using statistical modeling.

I use R to work with data, and Julia to construct statistical algorithms.

This quarter I started learning C and C++ from scratch -- so we're in the same boat in terms of learning new languages!

## What is R? 

- R is a programming language and environment for statistical computing and graphics. 

- R provides a wide variety of statistical and graphical techniques, and is highly extensible. 

- R is *not* a general-purpose programming language.

- R is free and open-source.

- R is used mainly by statisticians, both in the private sector (data scientists) and academia.

- R is used extensively within the economics profession due to its ease-of-use, flexibility, and reproducibility.

## What is RStudio?

RStudio is an IDE (Integrated Development Environment) for R. 

R is the language and RStudio helps us interact with R. 

Basically, if R is the engine, chassis, wheels, etc. of a car, then RStudio is its dashboard containing various buttons and monitors that, while "unneccessary", can make the driving experience easier and more effective.

You can still program in R without RStudio, but RStudio is useless without R. 

## Installation

Having R and RStudio on your laptop will allow you to work on problem sets and explore the functionality of R outside the lab. 

It is important that you **install R before you install R Studio.**

### Install R

In your web browser, go to [r-project.org](https://www.r-project.org) and then click [download R](https://cran.r-project.org/mirrors.html). 
At this point, you'll be directed to a page with a list of institutions that host the Comprehensive R Archive Network (CRAN). 
The idea is to pick an institution near you. Scroll down to USA and click the link for either [UC Berkeley](https://cran.cnr.berkeley.edu) or [OSU](https://ftp.osuosl.org/pub/cran/).

**Windows Instructions:** 
If you have a Windows machine, click `Download R for Windows` then `install R for the first time` then `Download R 4.1.3 for Windows`. 
To complete installation, run the `.exe` file you downloaded.

**Mac Instructions:** If you have a Mac, click `Download R for (Mac) OS X` then `R-4.1.3.pkg` under "latest release." 
To complete installation, run the `.pkg` file you downloaded.

**Linux Instructions:** If you run a Linux distro, note that installation instructions vary by distro. 
That said, you probably know what you're doing.

### Install RStudio

- In your web browser, go to [rstudio.com/products/rstudio/](https://rstudio.com/products/rstudio/), scroll down to R Studio Desktop, and then click [Download RStudio Desktop](https://rstudio.com/products/rstudio/download/) under "Open Source Edition."
Scroll down to "All Installers" and click the link that corresponds with your operating system. 
- To complete installation, run the installer you downloaded.

## R Basics

In this section we learn the absolute basic elements of R.

### R as a Calculator 

Addition:

```{r} 
2+2
```

Subtraction:

```{r} 
2-2
```

Multiplication:

```{r} 
2*2
```

Division:

```{r} 
2/2
```

Exponentiation:

```{r} 
2^2
```

(Natural) logarithms:

```{r} 
log(2)
```

### Variables

It is often useful to store information into objects/variables.
```{r}
a <- 2
b <- 3
c <- a * b + a^b - log(b)/a
```

The operator `<-` is referred to as the assignment operator.

Let's print variable `c` to see what it looks like:

```{r}
print(c) 
```

There exist many different classes of objects.
Let's see what class object `c` belongs to:
```{r}
class(c)
```

We can also create objects of class "character" (essentially text):
```{r}
sometext <- "Hello World" 
class(sometext) 
```

There are many other object classes -- we will run into new ones throughout the quarter.
A nice reference for the universe of standard object classes (data types) can be found [here](https://www.tutorialspoint.com/r/r_data_types.htm). 

### Functions

Functions take objects as inputs, and generate outputs that can be stored as new objects.

We've already run into one function -- `class`.

The `class` function takes any object as an input, and generates its class as an output.

Another great example of a function is `c`, which takes multiple inputs (of comparable dimensions) and combines them into a single vector:
```{r}
testvec <- c(a,b,c)
print(testvec)
```

We may now access the values of objects `a`, `b`, and `c` as elements of `testvec`:
```{r}
testvec[2]
```

We may also create out own functions:
```{r}
add <- function(arg1, arg2){
  arg1 + arg2
}
add(1,2)
```

### Packages

R functions come in packages. 

Each package contains a library of specialized functions.

Think of a package as a toolbox, and its functions as the actual tools -- a package simply keeps functions neatly grouped, organized, and labeled.

When you open a fresh R session in RStudio, a number of packages come pre-loaded. 
These include packages with common math and statistics functions and are known collectively as "base R." 

Base R is wonderful, but non-default packages offer a great deal of flexibility and functionality. 

**Install a package:** `install.packages("package.name.here")`

- Replace `package.name.here` with the name of the package you want to install.
- Make sure the name of the package is in quotes.
- You only have to install a package once.

**Load a package:** `library(package.name.here)`

- To use functions from a non-base package, you need to load the package at the beginning of each new R session.
- Quotes around the package name are no longer needed.

#### **`pacman`**

We will often need to load several packages in a single session. One way to do this is to execute `library(package.1)`, then `library(package.2)`, then `library(package.3)`, and so forth. A less cumbersome way to load multiple packages is to use the `p_load` function from the `pacman` package. 

- Use the instructions above to **install `pacman`**.
- Load the `pacman` package with `library(pacman)`.
- When you want to load additional packages, you can then execute `p_load(package.1, package.2, package.3)`.
- **Added bonus:** `p_load` first checks to see if the packages are installed. If they aren't, then it will install them for you.

```{r}
library(pacman)
p_load(ggplot2, pracma)
```

We have now loaded (and possibly installed) two packages called `ggplot2` and `pracma`. 
The former provides plotting capabilities, and the latter allows for easy random number generation.
We will use both of these packages below.

### R Scripts vs. The Console

To produce reproducible R code, it is best to use scripts. 
Open a new R script file with the `.R` extension by clicking `File` then `New File` then `R Script`. 
We will write our first script to generate a scatter plot using `ggplot2`.

- To execute *one line of your script*, click the line you want to run and then click `Run` at the upper-right corner of your R script. A quicker alternative is to click the line you want to run and then use the keyboard shortcut `Ctrl` `Enter`. 
- To execute *the entire script*, click `Source` at the upper-right corner of your R script or use the keyboard shortcut `Ctrl` `Alt` `R`.

**Aside:** It is useful to leave comments in your code to explain to your future self what your code is doing and why. You can leave a comment by typing a hash `#`:

```{r}
# This is a comment. R will ignore it.
```

### Data Frames

A data frame is a class of object often used to construct and store datasets. 

Let's starting by creating two variables, `var1` and `var2` that each contain 10 random numbers between 0 and 1.

```{r}
# Create `var1` and `var2`
var1 <- rand(10, 1)
var2 <- rand(10, 1)
```

Now if we would like to store `var1` and `var2` into a single table, we may create a data frame:

```{r}
# Create data frame containing `var1` and `var2`
df   <- data.frame(var1, var2)

# Print data frame
print(df)
```

### Plotting with `ggplot2`

To visualize relationships between variables stored in a data frame, you can make a scatter plot.

```{r}
# Plot `var2` on `var1`
ggplot(data = df, aes(x = var1, y = var2)) + 
  geom_point()
```

Now let's do the same thing using more data:

```{r}

# Create `var1` and `var2`
var1 <- rand(1000, 1)
var2 <- rand(1000, 1)

# Create data frame containing `var1` and `var2`
df   <- data.frame(var1, var2)

# Plot `var2` on `var1`
ggplot(data = df, aes(x = var1, y = var2)) + 
  geom_point()
```

