---
name: logic
---

---

## Regression models

Modeling is about reducing something [_really complicated_]{.hp} into something [_simple_]{.blue} that represents some part of the complicated reality.

- Try to tell stories that are easy to understand, and easy to learn from
- Model _toy versions_ of reality

. . .

Economists _often_ rely on [linear regression]{.hi} for statistical comparisons.

- Describes the relationship between a dependent (endogenous) variable and one or more explanatory (exogenous) variable(s)
- ["Linear"]{.note} is more flexible than you think


---

## Regression models

Regression analysis helps us make *all else equal* comparisons

. . . 

<br>

Running regressions provide correlative (_and even causal_) information between two variables

. . .

<br>

[Ex.]{.ex} By how much does $Y$ change when $X$ increases by one unit?

---

## Regression models

Modelling forces us to be explicit about the potential sources of selection bias

- Model the effect of $X$ on $Y$ while [controlling]{.hi} for [potential confounders]{.note} that may muddy the water
- Failure to account for sources of selection bias, leads to biased estimates.


<!-- ---

## Regression models {data-visibility="uncounted"}

Regression models forces us to be explicit about the potential sources of selection bias

- Model the effect of $X$ on $Y$ while [controlling]{.hi} [for potential confounders]{.hi} that may muddy the water
- Failure to account for sources of selection bias, leads to biased estimates. -->

. . .

[Ex.]{.ex} Not controlling for confounding variables, leads to [omitted-variable bias]{.note}, a close cousin of selection bias

. . .

- [Why?]{.note} [Omitted variables that correlate with our covariate of interest, hides within the model, distorting our results.]{.fragment}

---

## Returns to education

[Research Question:]{.note} By how much does an additional year of schooling increase future earnings?

- [Dependent variable:]{.blue} Earnings
- [Independent variable:]{.blue} An additional year of school

. . .

[Q.]{.note} How might education increase earnings?

. . .

[Q.]{.note} Why might a simple comparison between high and low educated not [isolate the economic returns to education]{.blue}?

---

## 

<!-- DAG SLIDE -->

More education ([X]{.hi}) increases lifetime earnings ([Y]{.hi}) 
<br>
<br>

```{r}
#| label: dag-ex-setup
#| echo: FALSE
#| include: FALSE



# The full DAG
dag_full = dagify(
  Y ~ X,
  Y ~ U,
  X ~ U,
  coords = tibble(
    name = c("Y", "X", "U"),
    x = c(1, 3, 2),
    y = c(2, 2, 1)
  )
)
# Convert to data.table
dag_dt = dag_full %>% fortify() %>% setDT()
# Add indicators for paths
dag_dt[, `:=`(
  path1 = (name == "X" & to == "Y") | (name == "Y"),
  path2 = (name == "X" & to == "U") | (name == "U" & to == "Y") | (name == "Y")
)]
# Shorten segments
mult = 0.15
dag_dt[, `:=`(
  xa = x + (xend-x) * (mult),
  ya = y + (yend-y) * (mult),
  xb = x + (xend-x) * (1-mult),
  yb = y + (yend-y) * (1-mult)
)]
```

```{r}
#| label: dag-ex-ovb-00
#| echo: FALSE
#| fig.height: 3
#| fig.width: 6
#| fig-align: center

# Filter data to include only D and Y nodes and the edge between them
dag_dt_filtered <- dag_dt[((name == "X" & to == "Y") | (name %in% c("X", "Y"))),]

# Plot the filtered DAG
ggplot(
  data = dag_dt_filtered,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = hi,
  shape = 21,
  stroke = 0.6
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = hi,
  size = 1.2,
  lineend = "round"
) +
geom_text(
  data = dag_dt_filtered[, .(name, x, y, xend = x, yend = y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = hi,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[, min(x)] * 0.95, dag_dt[, max(x)] * 1.05),
  ylim = c(dag_dt[, min(y)] * 0.8, dag_dt[, max(y)] * 1.1)
)

```

<br>
<br>


---

## {data-visibility="uncounted"}

More education ([X]{.hi}) increases lifetime earnings ([Y]{.hi}) along with a lot of other things ([U]{.hi}).

```{r}
#| label: dag-ex-ovb-01
#| echo: FALSE
#| fig.height: 3
#| fig.width: 6
#| fig-align: center

# Modify data to include only the desired edges
dag_dt_modified <- dag_dt[2:4]

# Plot the modified DAG
ggplot(
  data = dag_dt_modified,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = hi,
  shape = 21,
  stroke = 0.6
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = hi,
  size = 1.2,
  lineend = "round"
) +
geom_text(
  data = dag_dt_modified[, .(name, x, y, xend = x, yend = y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = hi,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[, min(x)] * 0.95, dag_dt[, max(x)] * 1.05),
  ylim = c(dag_dt[, min(y)] * 0.8, dag_dt[, max(y)] * 1.1)
)


```

<br>
<br>

---

## {data-visibility="uncounted"}

More education ([X]{.hi}) increases lifetime earnings ([Y]{.hi}) along with a lot of other things ([U]{.hi}). But a lot of other things ([U]{.hi}) also impact education ([X]{.hi}).


<!-- DAG SLIDE -->

```{r}
#| label: dag-ex-ovb-02
#| echo: FALSE
#| fig.height: 3
#| fig.width: 6
#| fig-align: center

# Plot the full DAG
ggplot(
  data = dag_dt,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = hi,
  shape = 21,
  stroke = 0.6
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = hi,
  size = 1.2,
  lineend = "round"
) +
geom_text(
  data = dag_dt[, .(name, x, y, xend = x, yend = y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = hi,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[, min(x)] * 0.95, dag_dt[, max(x)] * 1.05),
  ylim = c(dag_dt[, min(y)] * 0.8, dag_dt[, max(y)] * 1.1)
)
```

. . .

_Any unobserved variable_ that connects a [backdoor path]{.note} between education ([X]{.hi}) and earnings ([Y]{.hi}) is called a [confounder]{.hi}

---

## Returns to education

_How might we estimate the causal effect of an additional year of schooling on earnings?_

. . .

[Approach 1:]{.note} Compare average earnings of individuals who have 12 years of education to those with 16

- Prone to selection bias by variety of confounding variables

. . .

<!-- 
[Approach 2:]{.note} Use a matching estimator that compares the earnings of individuals the same admissions profiles.

- Cleaner comparison than a simple difference-in-means.
- Somewhat difficult to implement.
- Throws away data (inefficient).  

---

## Returns to education {data-visibility="uncounted"}

_How might we estimate the causal effect of private college on earnings?_ -->

[Approach 2:]{.note} Estimate a regression that compares the earnings of individuals with the same profiles

- _Try to_ [control]{.hi} for [confounders]{.note} by including them in the model
- Race, parental income, geography etc.

---

##

::: {.vertical-center}
But before taking on [confounders]{.note} and using regression to link causal relationships... [let's breakdown the anatomy of the [simple regression model]{.note}]{.fragment}
:::
