---
name: nonlinear
---

---

## Can We Do Better?

```{r}
#| include: false


reg_lin <- lm(lifeExp ~ gdpPercap, gapminder)
a_lin <- reg_lin$coefficients[1]
b_lin <- reg_lin$coefficients[2]
r2_lin <- summary(reg_lin)$r.squared
```


$$
(\widehat{\text{Life Expectancy})_i} = `r round(a_lin, 2)` + `r round(b_lin, 4)` \cdot \text{GDP}_i
$$

```{r}
#| echo: false
#| fig.height: 5.75

ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
geom_point(alpha = 0.75) +
scale_x_continuous("GDP per capita", label = scales::comma) +
ylab("Life Expectancy") +
stat_smooth(method = "lm", size = 1, color = hp, se = F) +
mytheme
```

---

## Nonlinear Relationships

Many economic relationships are **nonlinear**.

- *e.g.*, most production functions, profit, diminishing marginal utility, tax revenue as a function of the tax rate, *etc.*

. . .

[OLS is flexible]{.hi}

OLS can accommodate many, but not all, nonlinear relationships.

- Underlying model must be linear-in-parameters.
- Nonlinear transformations of variables are okay.
- Modeling some nonlinear relationships requires advanced estimation techniques, such as *maximum likelihood*^[Beyond the scope of this class.]

---

## Linearity

[Linear-in-parameters:]{.hi-green} [Parameters]{.green} enter model as a weighted sum, where the weights are functions of the variables.

- One of the assumptions required for the unbiasedness of OLS.

[Linear-in-variables:]{.hp} [Variables]{.purple} enter the model as a weighted sum, where the weights are functions of the parameters.

- Not required for the unbiasedness of OLS.

. . .

The standard linear regression model satisfies both properties:

$$Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + \dots + \beta_kX_{ki} + u_i$$

---

## Linearity

Which of the following is [linear-in-parameters]{.hi-green}, [linear-in-variables]{.hp}, or [niether]{.hi}?

[1.]{.note} $Y_i = \beta_0 + \beta_1X_{i} + \beta_2X_{i}^2 + \dots + \beta_kX_{i}^k + u_i$

[2.]{.note} $Y_i = \beta_0X_i^{\beta_1}v_i$

[3.]{.note} $Y_i = \beta_0 + \beta_1\beta_2X_{i} + u_i$

---

## Linearity {data-visibility="uncounted"}

Which of the following is [linear-in-parameters]{.hi-green}, [linear-in-variables]{.hp}, or [niether]{.hi}?

[1.]{.note} $\color{#A3BE8C}{Y_i = \beta_0 + \beta_1X_{i} + \beta_2X_{i}^2 + \dots + \beta_kX_{i}^k + u_i}$

[2.]{.note} $Y_i = \beta_0X_i^{\beta_1}v_i$

[3.]{.note} $Y_i = \beta_0 + \beta_1\beta_2X_{i} + u_i$

Model 1 is [linear-in-parameters]{.green}, but not [linear-in-variables]{.purple}. 

---

## Linearity {data-visibility="uncounted"}

Which of the following is [linear-in-parameters]{.hi-green}, [linear-in-variables]{.hp}, or [niether]{.hi}?

[1.]{.note} $\color{#A3BE8C}{Y_i = \beta_0 + \beta_1X_{i} + \beta_2X_{i}^2 + \dots + \beta_kX_{i}^k + u_i}$

[2.]{.note} $\color{#434C5E}{Y_i = \beta_0X_i^{\beta_1}v_i}$

[3.]{.note} $Y_i = \beta_0 + \beta_1\beta_2X_{i} + u_i$

Model 1 is [linear-in-parameters]{.green}, but not [linear-in-variables]{.purple}. 

Model 2 is [neither]{.hi}. 

---

## Linearity {data-visibility="uncounted"}

Which of the following is [linear-in-parameters]{.hi-green}, [linear-in-variables]{.hp}, or [niether]{.hi}?

[1.]{.note} $\color{#A3BE8C}{Y_i = \beta_0 + \beta_1X_{i} + \beta_2X_{i}^2 + \dots + \beta_kX_{i}^k + u_i}$

[2.]{.note} $\color{#434C5E}{Y_i = \beta_0X_i^{\beta_1}v_i}$

[3.]{.note} $\color{#B48EAD}{Y_i = \beta_0 + \beta_1\beta_2X_{i} + u_i}$

Model 1 is [linear-in-parameters]{.green}, but not [linear-in-variables]{.purple}. 

Model 2 is [neither]{.hi}. 

Model 3 is [linear-in-variables]{.purple}, but not linear-in-parameters.

---

## We're Going to Take Logs

The natural log is the inverse function for the exponential function: <br> $\quad \log(e^x) = x$ for $x>0$.
