---
name: quadratic
---

--- 

## Quadratic variables

Let's talk about a wage regression again. [Suppose we would like to estimate the effect of age on earnings. We estimate the following SLR:]{.fragment}

. . .

$$
\text{Wage}_i = \beta_0 + \beta_1 \text{Age}_i + u_i
$$

. . .

However, maybe we believe that $\text{Wage}_i$ and $\text{Age}_i$ have some nonlinear relationship[---the effect of an additional year of experience, when age is 27 vs age is 67, might be [different]{.hi}.]{.fragment} [So instead, we might estimate:]{.fragment}

. . .

$$
\text{Wage}_i = \beta_0 + \beta_1 \text{Age}_i + \beta_2 \text{Age}^2_i + u_i
$$

--- 

## Quadratic variables

In this model:

$$
\text{Wage}_i = \beta_0 + \beta_1 \text{Age}_i + \beta_2 \text{Age}^2_i + u_i
$$

the effect of $\text{Age}_i$ on $\text{Wage}_i$ would be:

. . .

$$
\frac{\partial \text{Wage}_i}{\partial \text{Age}_i} = \beta_1 + 2\beta_2 \text{Age}_i
$$

. . .



---

## Quadratic data

```{r}
#| echo: false
#| cache: true
#| fig-align: center
#| fig.width: 12
#| fig.height: 6.75

# Set seed
set.seed(1234)
# Sample size
n <- 1e3
# Generate data
quad_df <- tibble(
  x = runif(n, 21, 80),
  y = 30000 + 160*x - 1.5*x^2 + rnorm(n, sd = 500)
)
# regression
quad_reg <- lm(y ~ x + I(x^2), data = quad_df) 
b1 <- quad_reg$coefficients[2]
b2 <- quad_reg$coefficients[3]
# Plot
ggplot(data = quad_df, aes(x = x, y = y)) +
geom_point(size = 3, color = hi, alpha = 0.5) +
labs(
  title = "Relationship between age and income",
  subtitle = "Simulated data",
  x = "Age",
  y = "Earnings"
) +
mytheme
```

---

## Quadratic regression

[Regression Model]{.hi}

$$
Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + u_i
$$

. . .

[Interpretation]{.note}

Sign of $\beta_2$ indicates whether the relationship is convex ([+]{.mono}) or concave ([-]{.mono})

. . .

Sign of $\beta_1$? [🤷]{.fragment}

. . .

Partial derivative of $Y$ wrt. $X$ is the [marginal effect]{.hp} of $X$ on $Y$:

$$
\color{#B48EAD}{\dfrac{\partial Y}{\partial X} = \beta_1 + 2 \beta_2 X}
$$

- Effect of $X$ depends on the level of $X$

---

## Quadratic regression

```{r} 
lm(y ~ x + I(x^2), data = quad_df) %>% tidy()
```

[What is the marginal effect of]{.purple} $\color{#B48EAD}{X}$ [on]{.purple} $\color{#B48EAD}{Y}$[?]{.purple}
. . .
<br>
$\widehat{\dfrac{\partial \text{Y}}{\partial \text{X}} } = \hat{\beta}_1 + 2\hat{\beta}_2 X = `r round(b1, 2)` + `r round(b2*2, 2)`X$

---

## Quadratic regression

```{r}
lm(y ~ x + I(x^2), data = quad_df) %>% tidy()
```

[What is the marginal effect of]{.purple} $\color{#B48EAD}{X}$ [on]{.purple} $\color{#B48EAD}{Y}$ [when]{.purple} $\color{#B48EAD}{X=0}$[?]{.purple}

. . .

<br>

$\widehat{\dfrac{\partial \text{Y}}{\partial \text{X}} }\Bigg|_{\small \text{X}=0} = \hat{\beta}_1 = `r round(b1, 2)`$

---

## Quadratic regression

```{r}
lm(y ~ x + I(x^2), data = quad_df) %>% tidy()
```

[What is the marginal effect of]{.purple} $\color{#B48EAD}{X}$ [on]{.purple} $\color{#B48EAD}{Y}$ [when]{.purple} $\color{#B48EAD}{X=2}$[?]{.purple}

. . .

<br>

$\widehat{\dfrac{\partial \text{Y}}{\partial \text{X}} }\Bigg|_{\small \text{X}=2} = \hat{\beta}_1 + 2\hat{\beta}_2 \cdot (2) = `r round(b1, 2)` `r round(b2*2*2, 2)` = `r round(b1 + b2*2*2, 2)`$

---

## Quadratic regression

```{r}
lm(y ~ x + I(x^2), data = quad_df) %>% tidy()
```

[What is the marginal effect of]{.purple} $\color{#B48EAD}{X}$ [on]{.purple} $\color{#B48EAD}{Y}$ [when]{.purple} $\color{#B48EAD}{X=7}$[?]{.purple}
. . .
<br>
$\widehat{\dfrac{\partial \text{Y}}{\partial \text{X}} }\Bigg|_{\small \text{X}=7} = \hat{\beta}_1 + 2\hat{\beta}_2 \cdot (7) = `r round(b1, 2)` `r round(b2*2*7, 2)` = `r round(b1 + b2*2*7, 2)`$

---

```{r}
#| include: false

reg <- lm(y ~ x + I(x^2), data = quad_df)
margs <- margins::cplot(reg, x = "x", dx = "x", 
               what = "effect", draw = FALSE)
```

[Fitted regression line]{.hi}

```{r}
#| echo: false
#| cache: false
#| fig-align: center
#| fig.height: 7

quad_df %>%  
  ggplot(aes(x = x, y = y)) + 
  geom_point(color = hi, alpha = 0.5, size = 3) +
  stat_smooth(method = 'lm', formula = y ~ poly(x,2), se = F, linetype = 1, color = hp, size = 1) +
  xlab("X") + 
  ylab("Y") + 
  mytheme
```

---

[Marginal effect of $X$ on $Y$]{.hi}

```{r}
#| echo: false
#| cache: false
#| fig-align: center
#| fig.height: 7

margs %>%  
  ggplot(aes(x = xvals)) + 
  geom_line(aes(y = yvals), color = hp, size = 1) +
  geom_line(aes(y = upper), linetype = 2, color = hp, size = 1) +
  geom_line(aes(y = lower), linetype = 2, color = hp, size = 1) +
  geom_hline(yintercept = 0, linetype = 1, color = hi) +
  xlab("X") + 
  ylab("Marginal Effect") + 
  mytheme
```

---

## Quadratic regression

Where does the regression $\hat{Y_i} = \hat{\beta}_0 + \hat{\beta}_1 X_i + \hat{\beta}_2 X_i^2$ [turn?]{.hi}

<!-- - In other words, where is the peak (valley) of the fitted relationship? -->

. . .

[Step 1:]{.note} Take the derivative and set equal to zero.

$$
\widehat{\dfrac{\partial \text{Y}}{\partial \text{X}} } = \hat{\beta}_1 + 2\hat{\beta}_2 X = 0
$$

. . .

[Step 1:]{.note} Solve for $X$.

$$
X = -\dfrac{\hat{\beta}_1}{2\hat{\beta}_2}
$$

. . .

[Ex.]{.ex} Peak of previous regression occurs at $X = `r round(-b1/(2*b2), 2)`$.

# [Extra:]{.ex} Anscombe's Quartet {.inverse}

---

## Anscombe's Quartet

[Four *"identical"* regressions:]{.hi} Intercept [=]{.mono} 3, Slope [=]{.mono} 0.5, R^2^ [=]{.mono} 0.67

```{r}
#| echo: false
#| cache: true
#| fig-align: center
#| fig.height: 5.75

anscombe_m <- data.frame()

for(i in 1:4)
  anscombe_m <- rbind(anscombe_m, data.frame(set=i, x=anscombe[,i], y=anscombe[,i+4]))

anscombe_m %>% 
  ggplot(aes(x, y)) + geom_point(size=3, alpha = 0.8, color = hi) + 
  geom_smooth(method="lm", fill=NA, fullrange=TRUE, color = hp) + facet_wrap(~set, ncol=2) + 
  xlab("X") + 
  ylab("Y") + 
  mytheme
```
