---
name: ovb
---

---

## Omitted Variables

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center

venn_df <- tibble(
  x  = c( 0.0,   -0.5, -1.0),
  y  = c( 0.0,   -2.5, 2.0),
  r  = c( 1.9,    1.5, 1.3),
  l  = c( "Y", "X[1]", "X[2]"),
  xl = c( 0.0,   -0.5, -1.0),
  yl = c( 0.0,   -2.5, 2.2)
)

# Line types (order: x1, x2, x3, y)
venn_colors <- c(hii, hired, hi)
venn_lines <- c("solid", "dotted", "solid")
# Venn
ggplot(data = venn_df, aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(aes(linetype = l), alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
scale_linetype_manual(values = venn_lines) +
geom_text(aes(x = xl, y = yl, label = l), size = 9, family = "Fira Sans", parse = T) +
annotate(
  x = 1, y = 3,
  geom = "text", label = "No Bias", color = hp, size = 9, family = "Fira Sans", hjust = 0
) +
xlim(-3, 3) +
ylim(-4, 3.4) +
coord_equal()
```

---

## Omitted Variables

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center

# Colors (order: x1, x2, x3, y)
venn_lines <- c("solid", "dotted", "solid")
# Locations of circles
venn_df <- tibble(
  x  = c( 0.0,   -0.5, 1.5),
  y  = c( 0.0,   -2.5, -1.8),
  r  = c( 1.9,    1.5, 1.5),
  l  = c( "Y", "X[1]", "X[2]"),
  xl = c( 0.0,   -0.5, 1.6),
  yl = c( 0.0,   -2.5, -1.9)
)
# Venn
ggplot(data = venn_df, aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(aes(linetype = l), alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
scale_linetype_manual(values = venn_lines) +
geom_text(aes(x = xl, y = yl, label = l), size = 9, family = "Fira Sans", parse = T) +
annotate(
  x = 1, y = 3,
  geom = "text", label = "Bias", color = hp, size = 9, family = "Fira Sans", hjust = 0
) +
xlim(-3, 3) +
ylim(-4, 3.4) +
coord_equal()
```

---

## Omitted Variables

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center
#| escape: false

tab <- data.frame(
  v1 = c("Intercept", "", "log(Spend)", "", "Lunch", ""),
  v2 = rbind(
    c(-84.84, -1.52, ""),
    c("(18.57)", "(2.18)", "")
  ) %>% as.vector(),
  v3 = rbind(
    c(-6.34, 11.34, -0.47),
    c("(15.00)", "(1.77)", "(0.01)")
  ) %>% as.vector()
) %>% kable(
  escape = F,
  col.names = c("Explanatory variable", "1", "2"),
  align = c("l", rep("c", 3)),
  caption = "Math Score"
) %>%
row_spec(1:6, color = hp) %>%
row_spec(seq(2,6,2), color = "#c2bebe") %>%
row_spec(1:6, extra_css = "line-height: 110%;") %>%
kable_styling(font_size = 24) %>%  
column_spec(1, color = hi, italic = T)
tab %>% column_spec(2, bold = T)
```

Data from 1823 elementary schools in Michigan

- *Math Score*: average fourth grade state math test scores.
- *log(Spend)*: the natural logarithm of spending per pupil.
- *Lunch*: percentage of student eligible for free or reduced-price lunch.

---

## Omitted Variables

```{r}
#| echo: false
#| escape: false

tab %>% column_spec(3, bold = T)
```

Data from 1823 elementary schools in Michigan

- *Math Score*: average fourth grade state math test scores.
- *log(Spend)*: the natural logarithm of spending per pupil.
- *Lunch*: percentage of student eligible for free or reduced-price lunch.

---

## Omitted-Variable Bias

[Model 01:]{.hi} $Y_i = \beta_0 + \beta_1 X_{1i} + u_i$.

[Model 02]{.hii} $Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + v_i$

. . .

Estimating [Model 01]{.hi} (without $X_2$) yields [omitted-variable bias:]{.hp}

$$
\color{#B48EAD}{\text{Bias} = \beta_2 \frac{\mathop{\text{Cov}}(X_{1i}, X_{2i})}{\mathop{\text{Var}}(X_{1i})}}
$$

. . .

The sign of the bias depends on

1. The correlation between $X_2$ and $Y$, _i.e._, $\beta_2$.

2. The correlation between $X_1$ and $X_2$, _i.e._, $\mathop{\text{Cov}}(X_{1i}, X_{2i})$.


---

## Omitted variable bias

[OVB]{.hp} arises when we omit a variable, $X_k$ that

1. Affects the outcome variable $Y$, $\beta_k \neq 0$

2. Correlates with an explanatory variable $X_j$, $Cov(X_j, X_k) \neq 0$, 

Biases OLS estimator of $\beta_j$.

. . .

If we omit $X_k$, then the formula for the bias it creates in $\hat{\beta}_j$ is...

$$
\color{#B48EAD}{\text{Bias} = \beta_2 \frac{\mathop{\text{Cov}}(X_{1i}, X_{2i})}{\mathop{\text{Var}}(X_{1i})}}
$$


---

## Omitted variable bias

[Ex.]{.ex} Imagine a population model for the amount individual $i$ gets paid

$$
\text{Pay}_i = \beta_0 + \beta_1 \text{School}_i + \beta_2 \text{Male}_i + u_i
$$

where $\text{School}_i$ gives $i$'s years of schooling and $\text{Male}_i$ denotes an indicator variable for whether individual $i$ is male.

[Interpretation]{.note}

- $\beta_1$: returns to an additional year of schooling (*ceteris paribus*)
- $\beta_2$: premium for being male (*ceteris paribus*)

. . .

<br>

If $\beta_2 > 0$, then there is discrimination against women.

---

## Omitted variable bias

[Ex.]{.ex} From the population model

$$
\text{Pay}_i = \beta_0 + \beta_1 \text{School}_i + \beta_2 \text{Male}_i + u_i
$$

An analyst focuses on the relationship between pay and schooling, _i.e._,

$$
\text{Pay}_i = \beta_0 + \beta_1 \text{School}_i + \left(\beta_2 \text{Male}_i + u_i\right)
$$
$$
\text{Pay}_i = \beta_0 + \beta_1 \text{School}_i + \varepsilon_i
$$

where $\varepsilon_i = \beta_2 \text{Male}_i + u_i$.

---

## Omitted variable bias {data-visibility="uncounted"}

We assumed exogeniety to show that OLS is unbiased. 

[Even if $\mathop{\mathbb{E}}\left[ u | X \right] = 0$, it is not necessarily true that $\mathop{\mathbb{E}}\left[ \varepsilon | X \right] = 0$]{.fragment} 

. . .

- If $\beta_2 \neq 0$, then it is false

. . .

Specifically, if

$$
\mathop{\mathbb{E}}\left[ \varepsilon | \text{Male} = 1 \right] = \beta_2 + \mathop{\mathbb{E}}\left[ u | \text{Male} = 1 \right] \neq 0
$$

. . .

Then, OLS is [biased]{.hi}

---

## Omitted variable bias

Let's try to see this result graphically.

```{r}
#| include: false
#| cache: true

# Set seed
set.seed(12345)
# Sample size
n <- 1e3
# Parameters
beta0 <- 20; beta1 <- 0.5; beta2 <- 10
# Dataset
omit_df <- tibble(
  male = sample(x = c(F, T), size = n, replace = T),
  school = runif(n, 3, 9) - 3 * male,
  pay = beta0 + beta1 * school + beta2 * male + rnorm(n, sd = 7)
)
lm_bias <- lm(pay ~ school, data = omit_df)
bb0 <- lm_bias$coefficients[1] %>% round(1)
bb1 <- lm_bias$coefficients[2] %>% round(1)
lm_unbias <- lm(pay ~ school + male, data = omit_df)
bu0 <- lm_unbias$coefficients[1] %>% round(1)
bu1 <- lm_unbias$coefficients[2] %>% round(1)
bu2 <- lm_unbias$coefficients[3] %>% round(1)
```

The true population model:

$$
\text{Pay}_i = `r beta0` + `r beta1` \times \text{School}_i + `r beta2` \times \text{Male}_i + u_i
$$

The regression model that suffers from omitted-variable bias:

$$
\text{Pay}_i = \hat{\beta}_0 + \hat{\beta}_1 \times \text{School}_i + e_i
$$

Suppose that women, on average, receive more schooling than men.

---

## Omitted variable bias

True model: $\text{Pay}_i = `r beta0` + `r beta1` \times \text{School}_i + `r beta2` \times \text{Male}_i + u_i$

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center
#| escape: false

ggplot(data = omit_df, aes(x = school, y = pay)) +
geom_point(size = 2.5, color = hi, alpha = 0.4, shape = 16) +
xlab("Schooling") +
ylab("Pay") +
mytheme_s
```
---

## Omitted variable bias

Biased regression: $\widehat{\text{Pay}}_i = `r bb0` + `r bb1` \times \text{School}_i$

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center
#| escape: false

ggplot(data = omit_df, aes(x = school, y = pay)) +
geom_point(size = 2.5, color = hi, alpha = 0.4, shape = 16) +
geom_smooth(se = F, color = hii, method = lm) +
xlab("Schooling") +
ylab("Pay") +
mytheme_s
```

---

## Omitted variable bias

Recalling the omitted variable: Gender ([female]{.hp} and [male]{.hi})

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center
#| escape: false

ggplot(data = omit_df, aes(x = school, y = pay)) +
geom_point(size = 2.5, alpha = 0.8, aes(color = male, shape = male)) +
geom_line(stat = "smooth", color = hii, method = lm, alpha = 0.5, size = 1) +
xlab("Schooling") +
ylab("Pay") +
mytheme_s +
theme(
 legend.position = "none"
) +
scale_color_manual("", values = c(hp, hi), labels = c("Female", "Male")) +
scale_shape_manual("", values = c(16, 1), labels = c("Female", "Male"))
```

---

## Omitted variable bias

Recalling the omitted variable: Gender ([female]{.hp} and [male]{.hi})

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center
#| escape: false

ggplot(data = omit_df, aes(x = school, y = pay)) +
geom_point(size = 2.5, alpha = 0.8, aes(color = male, shape = male)) +
geom_line(stat = "smooth", color = hii, method = lm, alpha = 0.2, size = 1) +
geom_abline(
  intercept = lm_unbias$coefficients[1],
  slope = lm_unbias$coefficients[2],
  color = hp, size = 1
) +
geom_abline(
  intercept = lm_unbias$coefficients[1] + lm_unbias$coefficients[3],
  slope = lm_unbias$coefficients[2],
  color = hi, size = 1
) +
xlab("Schooling") +
ylab("Pay") +
mytheme_s +
theme(
 legend.position = "none"
) +
scale_color_manual("", values = c(hp, hi), labels = c("Female", "Male")) +
scale_shape_manual("", values = c(16, 1), labels = c("Female", "Male"))
```

---

## Omitted variable bias

Unbiased regression: $\widehat{\text{Pay}}_i = `r bu0` + `r bu1` \times \text{School}_i + `r bu2` \times \text{Male}_i$

```{r}
#| echo: false
#| fig.height: 5.75
#| fig-align: center
#| escape: false

ggplot(data = omit_df, aes(x = school, y = pay)) +
geom_point(size = 2.5, alpha = 0.8, aes(color = male, shape = male)) +
geom_line(stat = "smooth", color = hii, method = lm, alpha = 0.2, size = 1) +
geom_abline(
  intercept = lm_unbias$coefficients[1],
  slope = lm_unbias$coefficients[2],
  color = hp, size = 1
) +
geom_abline(
  intercept = lm_unbias$coefficients[1] + lm_unbias$coefficients[3],
  slope = lm_unbias$coefficients[2],
  color = hi, size = 1
) +
xlab("Schooling") +
ylab("Pay") +
mytheme_s +
theme(
 legend.position = "none"
) +
scale_color_manual("", values = c(hp, hi), labels = c("Female", "Male")) +
scale_shape_manual("", values = c(16, 1), labels = c("Female", "Male"))
```