---
name: interaction
---

---

## Motivation

Regression coefficients describe average effects. [But for whom does _on average_ mean?]{.fragment}

<br>

. . .

Averages can mask [heterogeneous effects]{.hp} that differ by group or by the level of another variable.

<br>

. . .

We can use [interaction terms]{.note} to model [heterogeneous effects]{.hp}[, accommodating complexity and nuance by going beyond *"the effect of* $X$ *on* $Y$ *is* $\beta_1$*."*]{.fragment}

---

## Interaction Terms

[Starting point:]{.hi} $Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + u_i$

:::: {.columns}
::: {.column width="50%"}
- $X_{1i}$ is the variable of interest
:::

::: {.column width="50%"}
- $X_{2i}$ is a control variable
:::

::::

. . .

[A richer model:]{.note} Interactions test whether $X_{2i}$ [*moderates*]{.purple} the effect of $X_{1i}$

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \beta_3 X_{1i} \cdot X_{2i} + u_i
$$

. . .

[Interpretation:]{.note} The partial derivative of $Y_i$ with respect to $X_{1i}$ is the [marginal effect]{.note} of $X_1$ on $Y_i$:

$$
\color{#81A1C1}{\dfrac{\partial Y}{\partial X_1} = \beta_1 + \beta_3 X_{2i}}
$$

. . .

The effect of $X_1$ depends on the level of $X_2$ 🤯


# [Ex.]{.ex} Differential returns to education {.inverse}

---

## [Ex.]{.ex} Differential returns to education

[Research question:]{.note} Do the returns to education vary by race?

. . .

Consider the interactive regression model:


\begin{align*}
  \text{Wage}_i = \beta_0 &+ \beta_1 \text{Education}_i + \beta_2 \text{Black}_i \\ 
                          &+ \beta_3 \text{Education}_i \times \text{Black}_i  + u_i
\end{align*}

. . .

_What is the marginal effect of an additional year of education?_

. . .

$$
\dfrac{\partial \text{Wage}}{\partial \text{Education}} = \beta_1 + \beta_3 \text{Black}_i  
$$

---

## [Ex.]{.ex} Differential returns to education

_What is the return to education for black workers?_

```{r}
#| echo: false
#| cache: true

reg = lm(wage ~ educ + black + educ:black, data = wage2)
# regression coefficients
a = reg$coefficients[2]
b = reg$coefficients[4]
```

```{r}
#| echo: true
#| cache: true

lm(wage ~ educ + black + educ:black, data = wage2) %>% tidy()
```

. . .

$$
\widehat{\left(\dfrac{\partial \text{Wage}}{\partial \text{Education}} \right)}\Bigg|_{\small \text{Black}=1} = \hat{\beta}_1 + \hat{\beta}_3 = `r round(a + b, 2)`
$$

---

## [Ex.]{.ex} Differential returns to education

_What is the return to education for non-black workers?_

```{r}
#| echo: true
#| cache: true

lm(wage ~ educ + black + educ:black, data = wage2) %>% tidy()
```

. . .

$$
\widehat{\left(\dfrac{\partial \text{Wage}}{\partial \text{Education}} \right)}\Bigg|_{\small \text{Black}=0} = \hat{\beta}_1  = `r round(a, 2)`
$$

---

## [Ex.]{.ex} Differential returns to education

[Q:]{.note} _Does the return to education differ by race?_

. . .

Conduct a [two-sided]{.hi} $t$-test of the [null hypothesis]{.note} that the interaction coefficient equals 0 at the 5% level.

. . .

<br>

```{r}
#| echo: true
#| cache: true

lm(wage ~ educ + black + educ:black, data = wage2) %>% tidy()
```

. . .

*p*-value [=]{.mono} 0.0499 < 0.05 [$\implies$ [reject null hypothesis]{.hi-red}.]{.fragment}

. . .

[A:]{.note} The return to education is significantly lower for black workers.

---

## [Ex.]{.ex} Differential returns to education

We can also test hypotheses about specific marginal effects.

- *e.g.,* H~0~: $\left(\dfrac{\partial \text{Wage}}{\partial \text{Education}} \right)\Bigg|_{\small \text{Black}=1} = 0$.
- Conduct a $t$ test or construct confidence intervals.

. . .

[Problem 1:]{.note} `lm()` output does not include $\hat{\text{SE}}$ for the marginal effects.

. . .

[Problem 2:]{.note} The formula for marginal effect standard errors includes covariances between coefficient estimates. The math is messy.^[Stay tuned]

. . .

[Solution:]{.note} Construct confidence intervals using the `margins` package.

---

## [Ex.]{.ex} Differential returns to education

The `margins` function provides standard errors and 95% confidence intervals for each marginal effect.

```{r}
#| echo: true
#| cache: true

pacman::p_load(margins)

reg = lm(wage ~ educ + black + educ:black, data = wage2)

margins(reg, at = list(black = 0:1)) %>% summary() %>% filter(factor == "educ")
```

---

## [Ex.]{.ex} Differential returns to education

We can use the `geom_pointrange()` option in `ggplot2` to plot the marginal effects with 95% confidence intervals.

```{r}
#| echo: false
#| cache: true
#| fig.width: 9
#| dev: svg

margs = summary(margins(reg, at = list(black = 0:1, south = 0:1))) %>% 
  filter(factor == "educ") %>% 
  mutate(Term = case_when(black == 1 ~ "Black Workers",
                          black == 0 ~ "Non-black Workers") %>% as.factor,
         Race = case_when(black == 1 ~ "Black", 
                          black == 0 ~ "Non-black") %>% as.factor)

margs %>% 
  ggplot() + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_pointrange(aes(x = Term, y = AME, ymin = lower, ymax = upper, color = Race), size = 1) +
  coord_flip() + mytheme_s + xlab("") + ylab("Marginal Effect of Education on Wages") +
  scale_color_manual("", values = c(hp, hi)) + 
  theme(
    legend.position = "none"
  ) 
```


